---
title: "Mini Data Analysis: Milestone 3"
date: "October 28, 2021"
author: "Katie Tjaden-McClement"
output: 
  github_document:
    toc: true
    toc_depth: 4
---


### Introduction

This is the 3rd and final milestone of the Mini Data Analysis project for STAT 545A. Milestone 1 focused on exploring datasets and coming up with research questions, milestone 2 used data wrangling and visualizations to begin to answer research questions. This milestone will continue the work from the first two by:

* Exploring special data types 
* Fitting models to answer research questions
* Reading and writing data in a robust manner

_Note: this Rmd was copied directly from the stat545 website as instructed, but instructions and other materials have been removed for conciseness and to maintain consistency of style within the MDA_

### Setup

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
library(lubridate) # for working with date data
library(broom) # produces tidy outputs of model objects

theme_set(theme_classic()) #sets theme for all plots as classic, removing backgrounds and gridlines
```

From Milestone 2, I chose two research questions:

1. How does having a root barrier relate to tree size as measured by diameter and height?
2. How does tree biomass (measured as a combination of number of trees and their size) vary across neighbourhoods?


## Exercise 1: Special Data Types

For this exercise, you'll be choosing two of the three tasks below.

But first, tasks 1 and 2 below ask you to modify a plot you made in a previous milestone. The plot you choose should involve plotting across at least three groups (whether by facetting, or using an aesthetic like colour). Place this plot below.

```{r M2 Tree_height_neighbourhood}
# Prepare trees dataset as in milestones 1 and 2:
trees <- vancouver_trees %>% 
  filter(diameter <= 75,
         diameter != 0)

trees %>% 
  mutate(neighbourhood_name = fct_rev(fct_infreq(neighbourhood_name))) %>% #reorder the factor by increasing frequency of neighbourhood
  group_by(neighbourhood_name) %>% 
  mutate(neighbourhood_count = n()) %>% # add a column to the tibble with tree count for each neighbouhood
  ggplot(aes(x = neighbourhood_name, y = height_range_id,
             colour = neighbourhood_count), # colour with a continuous scale according to # of trees in the neighbourhood
         alpha = 0.1) +
  geom_boxplot() +
  coord_flip() + # flips the x and y axes so that the neighbourhood names fit
  labs(y = "Tree height range", x = "Neighbourhood name",
       colour = "# of trees")
```

I have chosen to complete tasks 1 and 3.

**Task 1.** Produce a new plot that reorders a factor in your original plot, using the `forcats` package. Then, in a sentence or two, briefly explain why you chose this ordering.

I am reordering this plot to arrange the neighbourhoods by their mean tree height. This will allow a viewer to easily see which neightbourhoods have the tallest and shortest trees on average, because they will be at the top and the bottom of the plot, respectively.

```{r Tree height neighbourhood reordered}
trees %>% 
  mutate(neighbourhood_name = fct_reorder(neighbourhood_name, height_range_id, mean)) %>% #reorder the factor by mean tree height
  group_by(neighbourhood_name) %>% 
  mutate(neighbourhood_count = n()) %>% # add a column to the tibble with tree count for each neighbouhood
  ggplot(aes(x = neighbourhood_name, y = height_range_id,
             colour = neighbourhood_count), # colour with a continuous scale according to # of trees in the neighbourhood
         alpha = 0.1) +
  geom_boxplot() +
  coord_flip() + # flips the x and y axes so that the neighbourhood names fit
  labs(y = "Tree height range", x = "Neighbourhood name",
       colour = "# of trees")
```

This updated plot shows that Kitsilano and Shaugnessy have the tallest trees on average, while Renfrew-Collingwood and Oakridge have the shortest trees. Presenting the neighbourhoods in this order also highlights that there isn't a clear relationship between number of trees in a neighbourhood and their height, as some of the neighbourhoods with the most trees show up in the bottom half of this plot.

**Task 3.** If your data has some sort of time-based column like a date (but something more granular than just a year):
    1. Make a new column that uses a function from the `lubridate` or `tsibble` package to modify your original time-based column.
    2. Then, in a sentence or two, explain how your new column might be useful in exploring a research question.
    
```{r}
# First let's check out the class of the date_planted column and get a look at what the data looks like:
str(trees$date_planted) # "Date" class with YYYY-MM-DD date format

trees_month <- trees %>% 
  mutate(month_planted = month(date_planted, #extracts the month from the date planted data
                               label = TRUE), #labels month as ordered factor of written out month names, rather than numbers
                               abbr = TRUE) #abbreviates the month names
head(trees_month$month_planted)
```

This new month_planted variable could be used to investigate whether trees planted in certain months tend to have better outcomes e.g. larger diameters 15 years after planting, and if this is a species specific response. This line of questioning could lead to recommendations for best practices for what time of year to plant which species.

## Exercise 2: Modelling

### 2.0

Pick a research question, and pick a variable of interest that's relevant to the research question. Indicate these.

**Research Question**: How does having a root barrier relate to tree size as measured by diameter?

**Variable of interest**: Tree diameter

### 2.1

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen.

I will run a t-test to determine if the mean diameter of trees with root barriers is different from those without:

```{r Root barrier t-test}
root_barrier_ttest <- t.test(formula = diameter ~ root_barrier, data = trees)
root_barrier_ttest #print model object
```

### 2.2

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value. 

The _tidy_ function from the broom package converts a model output into a tibble that is easy to read and interpret. The value I want to highlight in this tibble is the t statistic, which is in the "statistic" column.
  
```{r}
tidy(root_barrier_ttest)
```

# Exercise 3: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there. 

First let's make sure the _here_ function is correctly identifying my project location:
```{r}
here::here() # yes, this is correct!
```

Now let's create an "Output" folder:
```{r}
dir.create(here::here("Output"))
```

### 3.1

Take a summary table that you made from Milestone 2 (Exercise 1.2), and write it as a csv file in your `output` folder. Use the `here::here()` function.

Table of summary statistics for trees with and without root barriers from milestone 2:
```{r}
q1_summary <- trees %>% 
  group_by(root_barrier) %>% 
  summarise(mean = mean(diameter, na.rm = T),
            range_lower = min(diameter),
            range_upper = max(diameter),
            median = median(diameter, na.rm = T),
            sd = sd(diameter),
            n = n())
q1_summary
```
Write as csv file in the Output folder:
```{r}
write_csv(q1_summary, here::here("Output", "q1_summary.csv"))
```

### 3.2

Write your model object from Exercise 2 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`. 

Write to RDS file in the output folder:
```{r}
saveRDS(root_barrier_ttest, here::here("Output", "root_barrier_ttest.RDS"))
```

Load back in:
```{r}
readRDS(here::here("Output", "root_barrier_ttest.RDS"))
```
