---
title: "Mini Data Analysis: Milestone 3"
date: "October 28, 2021"
author: "Katie Tjaden-McClement"
output: 
  github_document:
    toc: true
    toc_depth: 4
---


### Introduction

This is the 3rd and final milestone of the Mini Data Analysis project for STAT 545A. Milestone 1 focused on exploring datasets and coming up with research questions, milestone 2 used data wrangling and visualizations to begin to answer research questions. This milestone will continue the work from the first two by:

* Exploring special data types 
* Fitting models to answer research questions
* Reading and writing data in a robust manner

_Note: this Rmd was copied directly from the stat545 website as instructed, but instructions and other material has been removed for conciseness and to maintain consistency of style within the MDA_


### Setup

Begin by loading your data and the tidyverse package below: 

```{r, message = FALSE}
library(datateachr) # <- might contain the data you picked!
library(tidyverse)
library(lubridate) # for working with date data

theme_set(theme_classic()) #sets theme for all plots as classic, removing backgrounds and gridlines
```

From Milestone 2, I chose two research questions:

1. How does having a root barrier relate to tree size as measured by diameter and height?
2. How does tree biomass (measured as a combination of number of trees and their size) vary across neighbourhoods?


## Exercise 1: Special Data Types

For this exercise, you'll be choosing two of the three tasks below -- both tasks that you choose are worth 5 points each. 

But first, tasks 1 and 2 below ask you to modify a plot you made in a previous milestone. The plot you choose should involve plotting across at least three groups (whether by facetting, or using an aesthetic like colour). Place this plot below (you're allowed to modify the plot if you'd like). If you don't have such a plot, you'll need to make one. Place the code for your plot below.

```{r M2 Tree_height_neighbourhood}
# Prepare trees dataset as in milestones 1 and 2:
trees <- vancouver_trees %>% 
  filter(diameter <= 75,
         diameter != 0)

trees %>% 
  mutate(neighbourhood_name = fct_rev(fct_infreq(neighbourhood_name))) %>% #reorder the factor by increasing frequency of neighbourhood
  group_by(neighbourhood_name) %>% 
  mutate(neighbourhood_count = n()) %>% # add a column to the tibble with tree count for each neighbouhood
  ggplot(aes(x = neighbourhood_name, y = height_range_id,
             colour = neighbourhood_count), # colour with a continuous scale according to # of trees in the neighbourhood
         alpha = 0.1) +
  geom_boxplot() +
  coord_flip() + # flips the x and y axes so that the neighbourhood names fit
  labs(y = "Tree height range", x = "Neighbourhood name",
       colour = "# of trees")
```

Now, choose two of the following tasks.

**Task 1.** Produce a new plot that reorders a factor in your original plot, using the `forcats` package. Then, in a sentence or two, briefly explain why you chose this ordering.

I am reordering this plot to arrange the neighbourhoods by their mean tree height. This will allow a viewer to easily see which neightbourhoods have the tallest and shortest trees on average, because they will be at the top and the bottom of the plot, respectively.

```{r Tree height neighbourhood reordered}
trees %>% 
  mutate(neighbourhood_name = fct_reorder(neighbourhood_name, height_range_id, mean)) %>% #reorder the factor by mean tree height
  group_by(neighbourhood_name) %>% 
  mutate(neighbourhood_count = n()) %>% # add a column to the tibble with tree count for each neighbouhood
  ggplot(aes(x = neighbourhood_name, y = height_range_id,
             colour = neighbourhood_count), # colour with a continuous scale according to # of trees in the neighbourhood
         alpha = 0.1) +
  geom_boxplot() +
  coord_flip() + # flips the x and y axes so that the neighbourhood names fit
  labs(y = "Tree height range", x = "Neighbourhood name",
       colour = "# of trees")
```

This updated plot shows that Kitsalano and Shaugnessy have the tallest trees on average, while Renfrew-Collingwood and Oakridge have the shortest trees. Presenting the neighbourhoods in this order also highlights that there isn't a clear relationship between number of trees in a neighbourhood and their height, as some of the neighbourhoods with the most trees show up in the bottom half of this plot.

**Task 3.** If your data has some sort of time-based column like a date (but something more granular than just a year):
    1. Make a new column that uses a function from the `lubridate` or `tsibble` package to modify your original time-based column.
    2. Then, in a sentence or two, explain how your new column might be useful in exploring a research question.
    
```{r}
# First let's check out the class of the date_planted column and get a look at what the data looks like:
str(trees$date_planted) # "Date" class with YYYY-MM-DD date format

# We can use the function today() to return the current date
today()

trees_month <- trees %>% 
  mutate(month_planted = month(date_planted, #extracts the month from the date planted data
                               label = TRUE), #labels month as ordered factor of written out month names, rather than numbers
                               abbr = TRUE) #abbreviates the month names
head(trees_month$month_planted)
```

This new month_planted variable could be used to investigate whether trees planted in certain months tend to have better outcomes e.g. larger diameters 15 years after planting, and if this is a species specific response. This line of questioning could lead to recommendations for best practices for what time of year to plant which species.

## Exercise 2: Modelling

### 2.0 (no points)

Pick a research question, and pick a variable of interest (we'll call it "Y") that's relevant to the research question. Indicate these.

<!-------------------------- Start your work below ---------------------------->

**Research Question**: FILL_THIS_IN

**Variable of interest**: FILL_THIS_IN

<!----------------------------------------------------------------------------->

### 2.1 (5 points)

Fit a model or run a hypothesis test that provides insight on this variable with respect to the research question. Store the model object as a variable, and print its output to screen. We'll omit having to justify your choice, because we don't expect you to know about model specifics in STAT 545. 

- **Note**: It's OK if you don't know how these models/tests work. Here are some examples of things you can do here, but the sky's the limit. 
    - You could fit a model that makes predictions on Y using another variable, by using the `lm()` function.
    - You could test whether the mean of Y equals 0 using `t.test()`, or maybe the mean across two groups are different using `t.test()`, or maybe the mean across multiple groups are different using `anova()` (you may have to pivot your data for the latter two).
    - You could use `lm()` to test for significance of regression.

<!-------------------------- Start your work below ---------------------------->


<!----------------------------------------------------------------------------->

### 2.2 (5 points)

Produce something relevant from your fitted model: either predictions on Y, or a single value like a regression coefficient or a p-value. 

- Be sure to indicate in writing what you chose to produce. 
- Your code should either output a tibble (in which case you should indicate the column that contains the thing you're looking for), or the thing you're looking for itself.
- Obtain your results using the `broom` package if possible. If your model is not compatible with the broom function you're needing, then you can obtain your results by some other means, but first indicate which broom function is not compatible. 
  
<!-------------------------- Start your work below ---------------------------->


<!----------------------------------------------------------------------------->



# Exercise 3: Reading and writing data

Get set up for this exercise by making a folder called `output` in the top level of your project folder / repository. You'll be saving things there. 

## 3.1 (5 points)

Take a summary table that you made from Milestone 2 (Exercise 1.2), and write it as a csv file in your `output` folder. Use the `here::here()` function.

- **Robustness criteria**: You should be able to move your Mini Project repository / project folder to some other location on your computer, or move this very Rmd file to another location within your project repository / folder, and your code should still work.
- **Reproducibility criteria**: You should be able to delete the csv file, and remake it simply by knitting this Rmd file. 
  
<!-------------------------- Start your work below ---------------------------->


<!----------------------------------------------------------------------------->

## 3.2 (5 points)

Write your model object from Exercise 2 to an R binary file (an RDS), and load it again. Be sure to save the binary file in your `output` folder. Use the functions `saveRDS()` and `readRDS()`. 

- The same robustness and reproducibility criteria as in 3.1 apply here.

<!-------------------------- Start your work below ---------------------------->


<!----------------------------------------------------------------------------->

# Tidy Repository

Now that this is your last milestone, your entire project repository should be organized. Here are the criteria we're looking for.

## Main README (3 points)

There should be a file named `README.md` at the top level of your repository. Its contents should automatically appear when you visit the repository on GitHub. 

Minimum contents of the README file:

- In a sentence or two, explains what this repository is, so that future-you or someone else stumbling on your repository can be oriented to the repository.
- In a sentence or two (or more??), briefly explains how to engage with the repository. You can assume the person reading knows the material from STAT 545A. Basically, if a visitor to your repository wants to explore your project, what should they know?

Once you get in the habit of making README files, and seeing more README files in other projects, you'll wonder how you ever got by without them! They are tremendously helpful.

## File and Folder structure (3 points)

You should have at least four folders in the top level of your repository: one for each milestone, and one output folder. If there are any other folders, these are explained in the main README.

Each milestone document is contained in its respective folder, and nowhere else. 

Every level-1 folder (that is, the ones stored in the top level, like "Milestone1" and "output") has a `README` file, explaining in a sentence or two what is in the folder, in plain language (it's enough to say something like "This folder contains the source for Milestone 1").

## Output (2 points)

All output is recent and relevant: 

- All Rmd files have been `knit`ted to their output, and all data files saved from Exercise 3 above appear in the `output` folder. 
- All of these output files are up-to-date -- that is, they haven't fallen behind after the source (Rmd) files have been updated.
- There should be no relic output files. For example, if you were knitting an Rmd to html, but then changed the output to be only a markdown file, then the html file is a relic and should be deleted. 

Our recommendation: delete all output files, and re-knit each milestone's Rmd file, so that everything is up to date and relevant. 

PS: there's a way where you can run all project code using a single command, instead of clicking "knit" three times. More on this in STAT 545B!

## Error-free code (1 point)

This Milestone 3 document knits error-free. (We've already graded this aspect for Milestone 1 and 2)

## Tagged release (1 point)

You've tagged a release for Milestone 3. (We've already graded this aspect for Milestone 1 and 2)